Circumnuclear Media and Accretion Rates of Quiescent Supermassive Black Holes

### 2015-03-31 ms.tex
bdm2

In the presence of strong heating, one-dimensional steady-state flow
is characterized by an inflow-outflow structure, with a critical
radius known as the``stagnation radius" $\rs$ where the radial
velocity passes through zero.  Mass loss from stars interior to the
stagnation radius is accreted, while that outside $\rs$ is unbound in
an outflow from the nucleus.  The stagnation radius, rather than the
Bondi radius, thus controls the SMBH accretion rate (although we will
show that $\rs$ can in many cases lie within a factor of few of the
nominal Bondi radius).  When heating is sufficiently weak, however, no stagnation radius may exist at all, i.e. the entire ISM of the galaxy is flowing towards the SMBH.  A key goal of this work is to
assess under what conditions a stagnation radius exists, as this
question is intimately tied to that of cooling instability.
% Furthermore, in some cases we find that no stagnation
% radius exists at all; this formally implies that the entire of ISM of
% the galaxy is feeding the SMBH, potentially resulting in much higher
% accretion rate than would be predicted by simple Bondi accretion.

This paper is organized as followed.  In $\S\ref{sec:model}$ we
describe our model, including the sample of galaxies used in our
analysis ($\S\ref{sec:gal_model}$) and our numerical procedure for
calculating the steady-state hydrodynamic profile of the CNM
($\S\ref{sec:hydro}$).  In $\S\ref{sec:results}$ we describe some
analytic results, while in $\S\ref{sec:numerical}$ we describe our
numerical results.  In $\S\ref{sec:discussion}$ we discuss the
implications of our results, to topics including...  In
$\S\ref{sec:conclusions}$ we summarize our conclusions.  Table
\ref{table:definitions} provides the definitions of commonly used
variables.  Appendices \ref{app:rs} and \ref{app:be} provides useful
analytic results, while Appendix \ref{app:windheat} describes our
method for calculating stellar wind heating and mass input in greater
detail.

\begin{table*}
\begin{threeparttable}
%\centering
\begin{minipage}{18cm}
\caption{Definitions of commonly used variables}
\begin{tabular}{ll}
\hline
{Variable} & {Definition} \\
\hline
$M_{\bullet}$ & Black hole mass \\
$\tilde{v}_{w}$ & Total heating parameter, including minimum heating
rate from stellar and black hole velocity dispersion \\
$v_{w}$ & Total heating parameter, excluding heating rate from
velocity dispersion \\
$\sigma_0$ & Stellar velocity dispersion--assumed to be constant.\\
$\zeta$ & $\zeta=\sqrt{v_w^2+\sigma_0^2}/\sigma_0$\\
$r_{s}$ & Stagnation radius (where gas radial velocity goes to zero) \\
$r_{\rm soi}$ & radius of sphere of influence (eq.~[\ref{eq:rsoi}]) \\
$r_{\rm b}$ & outer break radius of stellar density profile \\ 
$r_{\rm Ia}$ & radius interior to which Ia SN are infrequent compared to the dynamical/inflow timescale (eq.~[\ref{eq:rIa}]) \\ 
$\rho_{\star}(r)$ & 3D radial stellar density profile \\
$\rho(r)$ & gas density of CNM \\
$M_{\star}(r)$ & total enclosed stellar mass inside radius $r$ \\
$M_{\rm enc}(r)$ & total enclosed mass inside radius $r$ (SMBH + stars) \\
$q(r)$ & mass source term due to stellar winds (eq.~[\ref{eq:q}]) \\
$\eta$ & parameter setting normalization of mass input from stellar winds (eq.~[\ref{eq:q}]) \\
$\tau_{\star}$ & age of stellar population, in case of a single burst of star formation \\
$\Gamma$ & power-law slope of radial stellar surface brightness
profile interior to the break radius \\
$\delta$ & power-law slope of the 3d stellar density profile inside of
the break radius. $\delta=\Gamma+1$.\\
$\dot{M}$ & mass accretion rate \\
$\dot{M}_{\rm Ia}$ & maximum mass accretion rate as limited by SN Ia blow-out (eq.~[\ref{eq:eddr_Ia}]) \\
$\dot{M}_{\rm C}$ & equilibrium accretion rate set by Compton heating (eq.~[\ref{eq:MdotC}]) \\
$\dot{M}_{\rm TI}$ & maximum accretion rate for thermally stable accretion (eq.~[\ref{eq:Mdotmax}]) \\
$\dot{q}_{\rm heat}/|\dot{q}_{\rm rad}|$ & ratio of external heating (stellar winds, SN Ia, MSPs) to radiative cooling (eq.~[\ref{eq:cooling2}]) \\
$t_{\rm h}$ & Hubble timescale \\
$\nu$ & gas density power-law slope at the stagnation radius
(eq.~[\ref{eq:densSlope}]) \\
\hline
\label{table:definitions}  
\end{tabular}
\begin{tablenotes}
\item{$^{(a)}FILL IN$}
\end{tablenotes}
\end{minipage}
\end{threeparttable}

@@

The term $\propto \vw^2 = \sigma(r)^2+v_{w}^2$ in the entropy equation
accounts for external heating sources (e.g.,
\citealt{ShcherbakovWong+:2014a}).  The first term approximates the
minimum amount of shock heating from stellar winds due to the random
movement of stars with respect to one another at the characteristic
velocity of $\sigma \approx \sqrt{\frac{3 G \Mbh}{(\Gamma+2)
    r}+\sigma_0^2}$, where $\sigma_0$ is a constant. The first term
corresponds to the velocity dispersion of a power law stellar density
profile with slope $-1-\Gamma$ in the presence of black hole. The
second term represents the observed constant velocity dispersion at
radii outside of the black hole sphere of influence
\citep{ChaeBernardi+:2014a}. For a given $\Mbh$, we take
$\sigma_0^2\approx 3 \sigma_{\Mbh}^2$, where $\sigma_{\Mbh}$ is the
velocity dispersion from the \citet{McConnellMa+:2011a} $\Mbh-\sigma$
relation.  This heating is present even in old stellar populations for
which mass and energy injection are dominated by AGB winds with low
velocities $\ll \sigma$.  The second term, $v_{w}^{2}$, parameterizes
all additional sources of heating, including faster main sequence or
Wolf-Rayet stellar winds, spin-down energy from millisecond pulsars,
supernovae, AGN feedback, etc.  As will be discussed in
$\S\ref{sec:heating}$, the value of $v_{w}$ will in general depend on
the SMBH mass and the age of the stellar population.  We further
assume that $v_w$ is constant with radius, i.e. that the volumetric
heating rate is proportional to the local stellar density.

>>>>

In the presence of strong heating, one-dimensional steady-state flow
is characterized by an inflow-outflow structure, with a critical
radius known as the``stagnation radius" $\rs$ where the radial
velocity passes through zero.  Mass loss from stars interior to the
stagnation radius is accreted, while that outside $\rs$ is unbound in
an outflow from the nucleus.  The stagnation radius, rather than the
Bondi radius, thus controls the SMBH accretion rate (although we will
show that $\rs$ can in many cases reside near the
nominal Bondi radius).  When heating is sufficiently weak, however, no stagnation radius may exist at all, i.e. the entire ISM of the galaxy is flowing towards the SMBH.  A goal of this work is to
assess under what conditions a stagnation radius exists, as this
question is intimately tied to that of cooling instability.
% Furthermore, in some cases we find that no stagnation
% radius exists at all; this formally implies that the entire of ISM of
% the galaxy is feeding the SMBH, potentially resulting in much higher
% accretion rate than would be predicted by simple Bondi accretion.

This paper is organized as followed.  In $\S\ref{sec:model}$ we
describe our model, including the sample of galaxies used in our
analysis ($\S\ref{sec:gal_model}$) and our numerical procedure for
calculating the steady-state hydrodynamic profile of the CNM
($\S\ref{sec:hydro}$).  In $\S\ref{sec:results}$ we describe some
analytic results, while in $\S\ref{sec:numerical}$ we describe our
numerical results.  In $\S\ref{sec:discussion}$ we discuss the
implications of our results, to topics including the nuclear X-ray luminosities of quiescent black holes, jetted TDEs, and the growth time of SMBHs.  In $\S\ref{sec:conclusions}$ we summarize our conclusions.  Table
\ref{table:definitions} provides the definitions of commonly used
variables.  Appendices \ref{app:rs} and \ref{app:be} provides useful
analytic results, while Appendix \ref{app:windheat} describes our
method for calculating stellar wind heating and mass input in greater
detail.

\begin{table*}
\begin{threeparttable}
%\centering
\begin{minipage}{18cm}
\caption{Definitions of commonly used variables}
\begin{tabular}{ll}
\hline
{Variable} & {Definition} \\
\hline
$M_{\bullet}$ & Black hole mass \\
$\tilde{v}_{w}$ & Total heating parameter, including minimum heating
rate from stellar and black hole velocity dispersion \\
$v_{w}$ & Total heating parameter, excluding heating rate from
velocity dispersion \\
$\sigma_0$ & Stellar velocity dispersion--assumed to be radially constant.\\
$\zeta$ & $\zeta=\sqrt{v_w^2+\sigma_0^2}/\sigma_0$\\
$r_{s}$ & Stagnation radius (where gas radial velocity goes to zero) \\
$r_{\rm soi}$ & radius of sphere of influence (eq.~[\ref{eq:rsoi}]) \\
$r_{\rm b}$ & outer break radius of stellar density profile \\ 
$r_{\rm Ia}$ & radius interior to which Ia SN are infrequent compared to the dynamical/inflow timescale (eq.~[\ref{eq:rIa}]) \\ 
$\rho_{\star}(r)$ & 3D radial stellar density profile \\
$\rho(r)$ & gas density of CNM \\
$M_{\star}(r)$ & total enclosed stellar mass inside radius $r$ \\
$M_{\rm enc}(r)$ & total enclosed mass inside radius $r$ (SMBH + stars) \\
$q(r)$ & mass source term due to stellar winds (eq.~[\ref{eq:q}]) \\
$\eta$ & parameter setting normalization of mass input from stellar winds (eq.~[\ref{eq:q}]) \\
$\tau_{\star}$ & age of stellar population, in case of a single burst of star formation \\
$\Gamma$ & power-law slope of radial stellar surface brightness
profile interior to the break radius \\
$\delta$ & power-law slope of the 3d stellar density profile inside of
the break radius. $\delta=\Gamma+1$.\\
$\dot{M}$ & mass accretion rate \\
$\dot{M}_{\rm Ia}$ & maximum mass accretion rate as limited by SN Ia blow-out (eq.~[\ref{eq:eddr_Ia}]) \\
$\dot{M}_{\rm C}$ & equilibrium accretion rate set by Compton heating (eq.~[\ref{eq:MdotC}]) \\
$\dot{M}_{\rm TI}$ & maximum accretion rate for thermally stable accretion (eq.~[\ref{eq:Mdotmax}]) \\
$\dot{q}_{\rm heat}/|\dot{q}_{\rm rad}|$ & ratio of external heating (stellar winds, SN Ia, MSPs) to radiative cooling (eq.~[\ref{eq:cooling2}]) \\
$t_{\rm h}$ & Hubble timescale \\
$\nu$ & gas density power-law slope at the stagnation radius
(eq.~[\ref{eq:densSlope}]) \\
\hline
\label{table:definitions}  
\end{tabular}
\begin{tablenotes}
\item{$^{(a)}FILL IN$}
\end{tablenotes}
\end{minipage}
\end{threeparttable}

@@

The term $\propto \vw^2 = \sigma(r)^2+v_{w}^2$ in the entropy equation
accounts for external heating sources (e.g.,
\citealt{ShcherbakovWong+:2014a}).  The first term approximates the
minimum amount of shock heating from stellar winds due to the random
movement of stars with respect to one another at the characteristic
velocity of 
\be\sigma \approx \sqrt{\frac{3 G \Mbh}{(\Gamma+2)
    r}+\sigma_0^2},
\ee
where the first term is the velocity dispersion of the power law stellar density profile in the SMBH potential.  The second term, $\sigma_0\approx 3 \sigma^2$ is a constant, motivated by the observed velocity dispersion outside of the SMBH sphere of influence \citep{ChaeBernardi+:2014a}, where $\sigma$ is the velocity dispersion in the $\Mbh-\sigma$
relation (eq.~[\ref{eq:Msigma}]).  The second component of the total heating, $v_{w}^{2}$, parameterizes all additional sources of energy input, including faster winds from young stars, millisecond pulsars, supernovae, AGN feedback, etc.  We assume that $v_w$ is constant with radius, i.e. that the volumetric heating rate is proportional to the local stellar density.  The value of $v_w$ will in general depend on the SMBH mass and the age of the stellar population ($\S\ref{sec:heating}$).
